---
title: "DACSS690C Homework 2"
author: "Patrick McGrath"
format: gfm
editor: visual
---

## Objective

The goal of this assignment is to examine the social network of Hollywood Actors.

```{r}
#| echo: false
#| warning: false
#| message: false

#load needed libraries and set random seed
library(igraph)
library(ggraph)
library(ggplot2)
library(reshape2)

#seed
set.seed(1111)
```

## Data

Data was downloaded from the provided link and then re-uploaded to dedicated Github for this assignment

```{r}
#| echo: false
dataGitLink='https://github.com/DACSS690C26/Homework_2/raw/main/hollywood.graphml'
actors=read_graph(dataGitLink,format='graphml')
#summary(actors)
```

```{r}
#plotting the base network while distinguishing Oscar winners
base=ggraph(graph = actors, layout = "stress")
base + geom_node_label(aes(label = name,
                           color=as.factor(wonOscar)),
                       repel = TRUE,show.legend = F) + 
  geom_edge_link(alpha=0.1) + 
  scale_color_manual(values = c('red','blue')) +
  labs(title = "Social Network of Hollywood Actors")
#blue = yes, red = no
```

### Calculate Centrality Measures

```{r}
eigen=eigen_centrality (actors)$vector


close = closeness(actors, normalized = TRUE)

betw <- betweenness(actors, normalized = TRUE)

DFCentrality=as.data.frame(cbind(eigen,close,betw),stringsAsFactors = F)
names(DFCentrality)=c('Eigenvector','Closeness','Betweenness')

DFCentrality$person=row.names(DFCentrality)
row.names(DFCentrality)=NULL

## see
head(DFCentrality)
```

```{r}
#plot closeness by betweenness with Eigenvector denoting size

ggplot(DFCentrality, aes(x=Betweenness, y=Closeness)) + 
  theme_classic() +
  geom_text(aes(label=person,size=Eigenvector),show.legend = T,alpha=0.5) 
```

```{r}
#| message : False


HubNodes=dplyr::slice_max(DFCentrality, order_by = Eigenvector, n = 3)$person
#HubNodes


NodeCount=length(V(actors))

V(actors)$label=''

for (index in seq(1:NodeCount)){
  currentName=V(actors)$name[index]
  if (currentName%in%HubNodes){
    V(actors)$label[index]=currentName
  }
}



base=ggraph(graph = actors, layout = "stress")
base  + geom_node_label(aes(label = label), 
                        repel = TRUE,
                        show.legend = F,
                        color='black') + 
  geom_edge_link(alpha=0.1) + 
  geom_node_point()
```

### Test for presence of Communities

Calculate the Ratio between the Actors transitivity and an ensemble of random networks. We found a ratio of 1.90, which being greater than 1, indicates that a community likely exists.

```{r}
# Generate an ensemble of 1000 rewired random networks 
set.seed(123)
replicates <- 1000  
random_transitivities <- replicate(replicates, {
  RandomNet <- rewire(actors, 
                      keeping_degseq(niter = gsize(actors) * 10))
  transitivity(RandomNet, type = "global")
})
mean_random_transitivities=mean(random_transitivities)

# Calculate your empirical transitivity
empirical_transitivity <- transitivity(actors, type = "global")

### Report
report_table <- data.frame(
  Metric = c("Actors transitivity", "Random-network mean", "Ratio"),
  Value  = round(c(empirical_transitivity,
                   mean_random_transitivities,
                   empirical_transitivity / mean_random_transitivities), 4)
)
knitr::kable(report_table)
```

### Analyze Possible Communities Via Various Algorithms

We tested the presence of communities within the social network via Louvain, Walk trap, Fast Greedy, Infomap, and Edge Betweenness, looking for the best algorithm.

```{r}
# Run all five community-detection algorithms
algos <- list(
  louvain          =  { set.seed(123); cluster_louvain(actors) },
   walktrap         = cluster_walktrap(actors),
   fast_greedy      = cluster_fast_greedy(actors),
   infomap          = cluster_infomap(actors),
  edge_betweenness = cluster_edge_betweenness(actors)
)

# Build a summary table: number of clusters, modularity (Q), and cluster sizes
summary_table <- data.frame(
  algorithm   = names(algos),
  n_clusters  = sapply(algos, length),
  modularity_Q  = sapply(algos, modularity),
  cluster_sizes = sapply(algos, function(cl) {
    paste(sort(sizes(cl), decreasing = TRUE), collapse = ", ")
  })
)

# Sort by modularity, descending, so the "best" partition (by this metric) is on top
summary_table <- summary_table[order(-summary_table$modularity_Q), ]
rownames(summary_table) <- NULL

print(summary_table)
```

Through examining the results of the various algorithms, we can see that they are relatively similar in modularity, with Walktrap, Infomap, and Edge Betweenness resulting in modularities of approximately 0.420, while Louvain and Fast Greedy resulted in modularity scores of 0.412. Both sets of modularity scores are moderate and indicative of string partitioning. The numbers of clusters present were found to be either 2 or 3, with Walktrap, Louvain, and Fast Greedy resulting in 3 clusters while Infomap and Betweenness resulted in 2. Cluster sizes were similar across algorithms with equal cluster sizes.

For our purposes, we can say that the Walktrap algorithm was the best method.

```{r}
for (name in names(algos)) {
  memb <- membership(algos[[name]])
  attr_name <- name
  actors <- set_vertex_attr(actors, attr_name,
                            index = names(memb),
                            # this is important for exporting
                            value = as.vector(memb))
}



#walktrap
base=ggraph(graph = actors, layout = "stress") +
  geom_edge_link(alpha=0.2)

base + geom_node_point(aes(color=as.factor(walktrap)), 
                       show.legend = T, size=4)+
  geom_node_text(aes(label = name), 
                 repel = TRUE, 
                 max.overlaps = 50, # avoid overlap
                 size = 2.5) +
  labs(title = "WalkTrap Algo Plot")
```

```{r}
#walktrap

walktrap_result <- algos$walktrap
memb <- membership(walktrap_result)
num_communities <- length(walktrap_result)

# comm_edges[i, j] will hold the density between community i and community j
# (diagonal = internal density, off-diagonal = cross-community density)
comm_edges <- matrix(0, nrow = num_communities, ncol = num_communities)

for (i in 1:num_communities) {
  for (j in 1:num_communities) {
    if (i == j) {
      # Internal density of a community -- edge_density() correctly uses
      # 2*edges / (n*(n-1)), no manual formula needed
      sub <- induced_subgraph(actors, which(memb == i))
      comm_edges[i, j] <- edge_density(sub)
    } else {
      # Cross density: edges crossing between community i and community j,
      # divided by every possible pair between the two groups
      nodes_i <- which(memb == i)
      nodes_j <- which(memb == j)
      el <- as_edgelist(actors, names = FALSE)
      boundary <- sum((el[,1] %in% nodes_i & el[,2] %in% nodes_j) |
                        (el[,1] %in% nodes_j & el[,2] %in% nodes_i))
      possible <- length(nodes_i) * length(nodes_j)
      comm_edges[i, j] <- boundary / possible
    }
  }
}

# Plot the density matrix as a heatmap; diagonal cells are internal density,
# off-diagonal cells show how "leaky" the boundary is between each pair
dimnames(comm_edges) <- list(1:num_communities, 1:num_communities)
df <- melt(comm_edges, varnames = c("Community_i", "Community_j"), value.name = "density")

ggplot(df, aes(x = factor(Community_j), y = factor(Community_i), fill = density)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.3f", density)), size = 4) +
  scale_fill_gradient(low = "#FFFFB2", high = "#BD0026", name = "Edge\ndensity") +
  scale_y_discrete(limits = rev) +
  coord_equal() +
  labs(title = "Edge density between communities", x = "Community", y = "Community") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

```{r}
#walktrap
V(actors)$degree = degree(actors) # first compute node degree and assign
ggraph(actors, layout = 'hive', 
       axis = walktrap,
       sort.by = degree
) +   
  geom_edge_hive(color='grey80') +   
  geom_axis_hive(aes(colour = as.factor(walktrap)), 
                 label = F) + 
  geom_node_point(aes(colour = as.factor(walktrap))) +
  geom_node_text(aes(label = name), 
                 repel = TRUE, 
                 max.overlaps = 50, # avoid overlap
                 size = 2.5) +
  coord_fixed()
```

```{r}
null_mean <- mean(random_transitivities)
null_sd   <- sd(random_transitivities)
z_score   <- (empirical_transitivity - null_mean) / null_sd
count_exceeding <- sum(random_transitivities >= empirical_transitivity)
p_value <- count_exceeding / replicates
test_table <- data.frame(
  Alpha    = c(0.05, 0.01),
  Z_score  = round(z_score, 3),
  P_value  = sprintf("%.3f (%d/%d)", p_value, count_exceeding, replicates),
  Decision = c(if (p_value < 0.05) "SIGNIFICANT" else "NOT significant",
               if (p_value < 0.01) "SIGNIFICANT" else "NOT significant")
)
knitr::kable(test_table)
```
